SSA is Freyd Categories

### 2023-11-01 abstract-freyd/paper.tex
Tap

Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure \cite{llvm}, GCC's GIMPLE IR, Cranelift \cite{cranelift}, and MLIR \cite{mlir} are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.

One of the interesting observations of the MLIR project is that SSA is very widely applicable, being useful for reasoning about everything from high-level machine learning operations to low-level GPU code, and even hardware design and simulation, as in the CIRCT project \cite{circt}. This hints that SSA is a natural syntax for some fundamental, hopefully compositional, structure, much like how the simply typed lambda calculus is a natural syntax for arbitrary Cartesian closed categories. In this talk, we would like to argue that SSA without control-flow is a natural syntax for \textit{Freyd categories}, which we then provide a graphical syntax for in terms of string diagrams as in \cite{promonad}. We then argue that SSA with \textit{terminating} control-flow (i.e., in which the basic blocks form a DAG) corresponds exactly to \textit{distributive} Freyd categories, while allowing general control-flow corresponds to equipping a distributive Freyd category with an \textit{Elgot structure}; we then give a string-diagrammatic graphical representation of such programs similar to traditional control-flow graphs.

We do all this by giving an inductive, type-theoretic presentation of SSA which we equip with a categorical semantics, proving substitution and the soundness of various basic optimizations and rewrites. Finally, we wish to give some examples of various models and their applications, highlighting how they appear in diverse areas of computer science including classical compiler development, weak memory, quantum circuits, and hardware design.

Time permitting, we also hope to explain how to generalize our IR with support for both substructural types and impure substitutions, which require us to move to the more general setting of \textit{distributive effectful categories}, and give a few applications of this more general system in separation logic and reasoning about nondeterminism.

>>>>

Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure \cite{llvm}, GCC's GIMPLE IR \cite{gcc}, Cranelift \cite{cranelift}, and MLIR \cite{mlir} are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.

One of the core observations of the MLIR project is that SSA is very widely applicable, being useful for representing everything from high-level machine learning operations to low-level GPU code, and even hardware designs, as in the CIRCT project \cite{circt}. This hints that SSA is a natural syntax for some fundamental compositional structure, much like how the simply typed lambda calculus is a natural syntax for Cartesian closed categories.

In this talk, we would like to argue that SSA without control-flow is a syntax for \textit{Freyd categories}, which we then provide a graphical syntax for in terms of string diagrams as in \cite{promonad}. We then argue that SSA with \textit{terminating} control-flow (i.e., in which the basic blocks form a DAG) corresponds exactly to \textit{distributive} Freyd categories, while SSA with general control-flow corresponds to equipping a distributive Freyd category with an \textit{Elgot structure}; we then give a string-diagrammatic graphical representation of such programs based on traditional control-flow graphs.

We do this by providing a type-theoretic presentation of SSA, which we call \isotopessa, which is designed to be amenable to inductive proofs and definitions. By using this formalization, we can equip SSA with a categorical semantics, which we show satisfies important properties such as semantic substitution and congruence. We then show how various optimizations, such as strength reduction, constant propagation, global value numbering, loop hoisting, and E-graph rewriting are compatible with our semantics. Finally, we give various models of our categorical semantics exhibiting interesting features, building up to a simple model of TSO-style weak memory based on \cite{sparky} which treats the interactions between threads in a compositional manner.

Time permitting, we also hope to explain how to generalize our IR with support for substructural types and impure substitutions, which require us to move to the more general setting of \textit{distributive effectful categories}, and give a few applications of this more general system relating to separation logic, quantum computing, and nondeterminism.
